---
title: "Historic Records"
author: "mikecp11@gmail.com"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Historic Records}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE, message = FALSE
  )
```

```{r setup, echo = FALSE}
library(nwmTools)
library(ggplot2)
```

As part of a collaborative research effort, the multi-decadal historic simulation products from the [NOAA National Water Model](https://water.noaa.gov/about/nwm) (versions 1.2, 2.0, and 2.1) have been reformatted and exposed through [CUAHSI](https://www.cuahsi.org/) at the [RENCI computing facility](https://renci.org/research/hydroshare/). This document outlines how to interact with this interface.

```{r}
knitr::kable(get_nwm_meta(c(1.2, 2.0, 2.1)))
```

## Data Extraction by ...

Reanalysis data can be found with the `readNWMdata` function. The API requires users to specify NHD Common Identifier(s) (COMID), or NWIS site ID(s) for time series extraction.

### NHD Catchment Identifier (COMID)

NWM reanalysis data can be extracted by NHD Catchment Identifier(s). The following code pulls the complete reanalysis data archive for COMID 17595383, a NHD reach near Goleta, California.

```{r}
library(nwmTools)

system.time({
  flows = readNWMdata(comid = 17595383)
})

head(flows)
dim(flows)
```

### USGS NWIS Site Number

Alternatively, data can be extracted using USGS NWIS site ID(s).

```{r}
system.time({
  flows = readNWMdata(siteID  = 11120000)
})

head(flows)
dim(flows)
```

## Query modification by ...

The default behavior is to extract the complete, hourly time series in UTC, for the most current model version. Parameters allow data to be modified by data range, timezone and model version as follows: 

### date range

By default, the API extracts the complete flow record for each ID. This range can constrained by providing a `startDate` and/or `endDate` in the form of ("YYYY-MM-DD"). 

```{r}
system.time({
  utc <- readNWMdata(comid = 101, 
                     startDate = "2010-01-01", 
                     endDate   = "2010-01-31")
})

# Check time dimensions
dim(utc)
range(utc$dateTime)
head(utc)
```

### time-zone

NWM output is recorded in UTC. Instead, users can specify a timezone. For our COMID in Goleta, California we can request data in `US/Pacific` time. Valid times zone can be found with `OlsonNames()`.

```{r}
pst = readNWMdata(comid = 101, 
                  startDate = "2010-01-01", 
                  endDate = "2010-01-31", 
                  tz = "US/Pacific")

# Check time dimensions
dim(pst)
range(pst$dateTime)
head(pst)
```

Finally, we can plot the UTC return against the PST return:

```{r, echo = FALSE}
ggplot() +
  geom_line(data = utc, aes(x = dateTime, 
                            y = flow_cms_v2.1, 
                            color = "UTC")) +
  geom_line(data = pst, aes(x = utc$dateTime, 
                            y = flow_cms_v2.1, 
                            color = "PST")) +
  theme_bw() +
  labs(y = 'Flows', x = "DateTime", title = "Time Zone Comparison", color = "Timezone") + 
  theme(legend.position = "bottom")
```

### Model versioning

The reanalysis product for versions 1.2, 2.0, and 2.1 of the NWM are available. By default the API extracts data from version 2.1, but all versions can be requested.

```{r}
pst_vs = readNWMdata(comid = 101, 
                     startDate = "2010-01-01", 
                     endDate   = "2010-01-31",
                     tz        = "US/Pacific",
                     version =  c(1.2, 2.0, 2.1))
```

```{r, echo = FALSE}
ggplot(data = pst_vs) + 
  geom_line(aes(x = dateTime, y = flow_cms_v1.2, color = "v1.2")) + 
  geom_line(aes(x = dateTime, y = flow_cms_v2,   color = "v2.0")) +
  geom_line(aes(x = dateTime, y = flow_cms_v2.1, color = "v2.1")) +
  labs(y = 'Flows', x = "DateTime", title = "Model Comparison", 
       color = "Model") + 
  theme_bw() +
  theme(legend.position = "bottom")
```

### model version vs obs

In addition to extracting records from mulitple model runs, data can be appended from observation in the case a USGS gage ID is supplied:

```{r}
system.time({
 data = readNWMdata(siteID = '01017000', 
            startDate = "2015-01-01",
            endDate = "2016-01-01",
            version = c(1.2, 2.0, 2.1),
            addObs = TRUE)
})
```

```{r, echo = FALSE}
ggplot(data = data) + 
  geom_line(aes(x = dateTime, y = flow_cms_v1.2, color = "v1.2")) + 
  geom_line(aes(x = dateTime, y = flow_cms_v2,   color = "v2.0")) +
  geom_line(aes(x = dateTime, y = flow_cms_v2.1, color = "v2.1")) +
  geom_line(aes(x = dateTime, y = flow_cms_nwis, color = "nwis")) +
  labs(y = 'Flows', x = "DateTime", title = "Model Comparison vs Observation", 
       color = "Model") + 
  theme_bw() +
  theme(legend.position = "bottom")
```

In section 2, we will look at how we can facilitate COMID and USGS NWIS discovery. In section 3 we will look at some techniques to handle operational data, and in section 4, we will look at time aggregations and splitting.

