---
title: "Feature Discovery"
author: "mikecp11@gmail.com"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Feature Discovery}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE, message = FALSE
  )
```

NWM channel output is indexed by NHD COMID, or NWIS site ID. Identifying the relevant IDs is not always easy. Tools exist in `dataRetrieval` and `nhdplusTools` to aid in (1) location and (2) area based feature discovery.

### Location based search

NHD COMIDs can be determined from point locations as the NHD catchment containing the requested point. Here we find the COMID associated with a specific lat/lon.

```{r}
library(nwmTools)
library(AOI)
library(dataRetrieval)
library(nhdplusTools)
library(ggplot2)

# Find COMID
(findNLDI(location = c(-119.8276, 34.43583)))
```

### Area Search

`nhdplusTools::get_nhdplus(...)` can accept polygon features as a search domain and will return all NHD flowlines intersecting the input area. From theses, COMIDs can be extracted.

#### Define Area...

```{r}
# Fort Collins footprint
(AOI = aoi_get("Fort Collins"))
```

#### Pull COMIDS for the AOI.
```{r}
# Find COMIDS
ids = get_nhdplus(AOI)$comid
head(ids)
length(ids)
```

### Finding NWIS gages and pull siteIDs...

Analogous functionality for finding USGS NWIS site IDs is provided with `nhdplusTools::get_nwis`. This function only returns NWIS sites that record streamflow (USGS parameter code '00060') and are collocated with an NHD catchment represented in the NWM.

```{r}
ids = get_nwis(AOI)$site_no
head(ids)
length(ids)
```

## Geometric Discovery

The NHDPlusV2 data model loosely conforms to the HY_Features Conceptual Model with a mapping shared [here](https://github.com/opengeospatial/HY_Features/wiki/HY_Features-NHDPlus-Mapping). `nhdplusTools` provides the ability to retrieve catchment-divide, flowline, and outlet representations of a hydrologic landscape, for an AOI.

```{r}
# Return catchments
fc = get_nhdplus(AOI, realization = "all")
str(fc, max.level = 1)
```

```{r, echo = FALSE}
ggplot() +
  geom_sf(data = fc$catchment, aes(fill = "Catchment"), color = "black") +
  geom_sf(data = fc$flowline,  aes(color = 'Flowline'), show.legend  = "line") +
  geom_sf(data = fc$outlet, aes(color = 'Outlet'), show.legend  = "point") + 
  scale_fill_manual(values = c("Catchment" = "gray99"), 
                    guide = guide_legend(override.aes = list(linetype = 1, border = "solid", shape = NA))) +
  scale_colour_manual(values = c("Flowline" = "blue","Outlet" = "red"), 
                      guide = guide_legend(override.aes = list(linetype = c("solid", "blank"), shape = c(NA,16)))) + 
  theme_minimal() +
  labs(fill = "", color = "") + 
  theme_void() +
  theme(legend.position = 'bottom')
```

## Putting it all together

Lets look at one integrative example. The aim is to identify a self-contained watershed and extract reanalysis records along the mainstem for 2015. The `dataRetrival` Network Linked Data Index can be used to define a starting point, and traverse the hydrographic network to find the upstream (or downstream) geometries and indexed elements. Once the COMIDs are identified, `nwmTools` can be used to extract the relevant streamflow forecasts.

```{r}
loc = findNLDI(nwis = '05428500', 
               nav = c("UT", "UM"), 
               find = c("flowline", "basin"))

# Find 2015 Flows along the Mainstem
nldi_flows <- readNWMdata(comid = loc$UM_flowlines$nhdplus_comid, 
                          startDate = "2015-01-01", 
                          endDate   = "2015-12-31")
```

```{r, echo = FALSE, fig.show="hold", fig.width=7.5, fig.height=4}
um_outlets = get_node(loc$UM_flowlines)
um_outlets$order = factor(nrow(um_outlets):1)

# Make a map
ggplot() +
  geom_sf(data = loc$basin, col = 'gray90') +
  geom_sf(data = loc$UT_flowlines, col = 'lightblue') + 
  geom_sf(data = loc$UM_flowlines, col = 'blue', size = .5) + 
  geom_sf(data = um_outlets, aes(color = order), size = 1.25) + 
  scale_color_viridis_d() +
  theme_void() + theme(legend.position="none")


# Plots the flows
ggplot(data = nldi_flows, aes(x = dateTime, y = flow_cms_v2.1, color = comid)) +
  geom_line(size = .25) + 
  scale_color_viridis_d() +
  labs(y = "Qsim (cms)",
       x = paste0("DateTime (", base::format(nldi_flows$dateTime[1], format="%Z"), ")"), 
       title = paste("Flow Records")) +
  theme(legend.position = 'NA')
```
