---
title: "Time Aggregation{s}"
author: "mikecp11@gmail.com"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time Aggregation{s}}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE, message = FALSE
)
```

```{r setup, echo = FALSE}
library(ggplot2)
library(dplyr)
library(nwmTools)
```

Raw historic data comes in hourly time intervals. For many analyses, different time aggregates are needed (e.g. by year, by-month ...). To ease repetative tasks, the `nwmTools` provides a `split_time` function and a family of `aggregate_*` functions.


## Split-time

`split_time` in breaks the hourly `dateTime` into base temporal units. In addition to the basic year, month, day, and hour, the season, water year, and Day-of-Water Year (DOWY) are added: 

```{r}
# Get 5-years of data and split the time attribute
st = readNWMdata(comid = 101, 
                 startDate = "2005-01-01", 
                 endDate = "2009-12-31") |> 
  split_time("dateTime")

glimpse(st)
```
Using the split time attributes, we can plot the hourly flow records by julien day, grouped by water-year:

```{r}
ggplot(data = st, aes(x = DOWY, y = flow_cms_v2.1, color = flow_cms_v2.1)) + 
  geom_line(size = 1) +
  facet_grid(wy~.) +
  labs(y= "Daily Flow (cms)", x= "Day of Water Year", 
       title="Daily Discharge for COMID 101") +
  scale_color_viridis_c() +
  theme_minimal()
```

## Aggregation

Often you might want to split and summarize your data, for example "_average monthly flow_" or "_median annual flow_". 

For these tasks, a family of aggregation methods allow users to define an temporal unit via the function name, and pass summarizing function(s) as parameters. Function names follow the pattern of `aggregate_*` where `*` represents the common date (and hydro-specific) symbols seen below.

| Symbol          | Aggregate          |
| --------------- |:------------------:|
| y               | year               |
| m               | month              |
| d               | day of moth        |
| doy             | day of year        |
| j               | Julian day         |
| s               | season             |
| wy              | water year         |
| dowy            | day of water year  |


These symbols can be combined to provide useful, common aggregation patterns, `r length(grep("aggregate",ls("package:nwmTools")))` of these are included the package (some are shown below):

| Aggregate Unit Symbol         |       Description    | 
| ----------------| :---------------------------------:| 
| *_record        | Entire Record                      | 
| *_y             | Year                               | 
| *_m             | Month                              | 
| *_j             | Julian Day                         | 
| *_s             | season                             | 
| *_wy            | Water Year                         | 
| *_ym            | Year and Month                     | 
| *_yj            | Year and Julian day                | 
| *_ymd           | Day of the Year                    | 
| *_ys            | Year and Season                    | 
| *_wym           | Water Year and Month               | 
| *_wymd          | Julian Day of the Water Year       | 
| *_wys           | Water Year and Season              |
| *_dowy          | Day of Water Year                  |

### Examples

First lets grab some data for a COMID near Baton Rouge, LA along the Mississippi.

```{r}
flows = readNWMdata(comid = 101)
glimpse(flows)
```

### Single function

Using the flow data grabbed above, we might be interested in seeing the monthly mean flow rates. We can do this by passing the flow records to `aggregate_m` (m = month) and using `mean` as the function:

```{r}
# Aggregate hourly flows to monthly averages by year
monthly = aggregate_m(flows, fun = mean)
glimpse(monthly)

ggplot(data = monthly) + 
  geom_col(aes(x = factor(month), y = flow_cms_v2.1)) +
  labs(x = "Month", y = "Q (cms)", 
       title = 'Monthly Average') + 
  theme_bw()
```

Alternatively we might be interested in the monthly variability in each year. We can do this by passing the flow records to `aggregate_ym` (ym = year,month) and using `mean` as the summarizing function:

```{r}
# Aggregate hourly flows to monthly averages by year
ym = aggregate_ym(flows, fun = mean)
glimpse(ym)

ggplot(data = ym, aes(x = factor(month), y = flow_cms_v2.1)) + 
  geom_boxplot(alpha = .5, outlier.shape = NA) +
  labs(x = "Month", y = "Q (cms)", 
       title = 'By Month Variability') + 
  theme_bw()
```

### Multiple functions

So far we have only looked at passing `mean` to `aggregate_*` but multiple functions can also be passed as a vector. The following returns the seasonal (s) mean and standard deviation.

```{r}
# Aggregate by season
seasons = aggregate_s(flows, fun = c('mean', 'sd'))
glimpse(seasons)

ggplot(data = seasons, aes(x = season, y = flow_cms_v2.1_mean)) +
  geom_errorbar(aes(ymin=flow_cms_v2.1_mean - flow_cms_v2.1_sd, ymax=flow_cms_v2.1_mean + flow_cms_v2.1_sd, width = .25)) +
  geom_point(color = "darkred", size = 2) +
  labs(x  = "Season", y = "Q(cms)", 
       title = "Seasonal mean +- sd")
```


### Custom Functions

Equally important, you are not limited to base R functions. Instead you can pass any function to `fun` that works over a vector of streamflow elements. In the code below we ask for a number of percentiles along with some other summary statistics for every junlien day of the year:

```{r}
# Aggregate by Julien Day
jul = aggregate_j(flows, fun = c(
                          n05 = function(x){quantile(x,.05)},
                          n25 = function(x){quantile(x,.25)},
                          n75 = function(x){quantile(x,.75)},
                          n95 = function(x){quantile(x,.95)},
                          median = median,
                          mean   = mean,
                          min    = min,
                          max    = max))

glimpse(jul)


ggplot(data = jul, aes(x = julian)) +
  geom_ribbon(aes(ymin=flow_cms_v2.1_min, ymax=flow_cms_v2.1_max), fill="#B2DFEE") +
  geom_ribbon(aes(ymin=flow_cms_v2.1_n05, ymax=flow_cms_v2.1_n95), fill="#9AC0CD") +
  geom_ribbon(aes(ymin=flow_cms_v2.1_n25, ymax=flow_cms_v2.1_n75), fill="#68838B") +
  geom_line(aes(y = flow_cms_v2.1_mean), col = "#104E8B" ) +
  geom_line(aes(y = flow_cms_v2.1_median), col = "#AFEEEE") + 
  theme_bw() +
  labs(x = "Day of Year", y = "Q (cms)", 
       title = "Flows @ COMDI 101...",
       subtitle = "26 years of record")
```


